sdmTMB statistical model description

### 2021-12-09 doc/appendix-model.Rmd
censored-poisson?
Add to the model appendix

* Bold subscript $\s$ since x, y

@@

* $\phi$ for all 'dispersion' parameters for consistency with code

* Family titles link to TMB or sdmTMB source code

* Attempt to link math symbols to argument options in model and code symbols

```{r, results='asis'}
symbols <- tibble::tribble(
  ~Symbol, ~Description,
  "$\\phi$", "A dispersion parameter for a distribution",
  "$\\mu$", "Mean",
  "$y$", "Observed response data",
  "$f$", "Link function",
  "$\\omega_{\\s}$", "Spatial random field at point $\\s$",
  "$\\zeta_{\\s, t}$", "Spatially varying coefficient random field at point $\\s$ and time $t$",
  "$\\epsilon_{\\s, t}$", "Spatiotemporal random field",
  "$\\boldsymbol{\\Sigma}_\\omega$", "Spatial random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\epsilon$", "Spatiotemporal random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\zeta$", "Spatially varying coefficient random field covariance matrix",
  "$\\boldsymbol{\\beta}$", "Parameter vector",
  "$\\boldsymbol{X}$", "Model predictor matrix",
  "$\\alpha_g$", "IID random intercept deviations",
  "$\\sigma_\\alpha^2$", "IID random intercept variance",
  "$\\delta_{\\s,t}$", "AR1 or random walk spatiotemporal deviations",
  "$\\rho$", "Correlation between random fields in subsequent time steps"

)
knitr::kable(symbols, caption = "Symbol notation and descriptions", escape = FALSE, format = "latex", booktabs = TRUE)
```

@@

* $y$ represents the response data;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

@@

## Main effects:

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} \ldots \right)
\end{aligned}
$$

Within `sdmTMB()`,

* $\boldsymbol{X} \boldsymbol{\beta}$ is defined by the `formula` argument and
  represents the fixed effect model matrix and a corresponding vector of
  coefficients.

@@

Smoothers in sdmTMB are generally implemented with the same formula syntax familiar to `mgcv` users fitting GAMs (generalized additive models). Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`. Within these smooths, the same syntax commonly used in mgcv::s() can be applied, e.g. 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; the basis function dimensions may be specified, e.g. `+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, `+ s(month, bs = "cc", k = 12)`.

While `mgcv` can fit unpenalized (e.g. B-splines, and others) or penalized splines (P-splines), sdmTMB only uses penalized splines. 
The penalized splines are constructed in sdmTMB using the function `mgcv::smooth2random()`, which transforms splines into random effects (and associated design matrices) that are estimable in a mixed effects modeling framework. This is a similar approach implemented in `gamm4` and `brms`, among other packages.

@@

The linear breakpoint or hockey stick model can be used to describe threshold or asymptotic responses. This function consists of two pieces, so that for $x < b_{1}$

$$s(x) = x \cdot b_{0}$$
and for $x > b_{1}$

$$s(x) = b_{1} \cdot b_{0}$$
In both cases, $b_{0}$ represents the slope of the function up to a threshold, and the product $b_{1} \cdot b_{0}$ represents the value at the asymptote. No constraints are placed on parameters $b_{0}$ or $b_{1}$.

@@

The form is

@@

where $\psi$ is a scaling parameter (controlling the height of the y-axis for
the response, and is unconstrained), $\tau$ is an intercept, $s50$ is
a parameter controlling the point at which the function reaches 50% of the
maximum ($\psi$), and $s95$ is a parameter controlling the point at which the
function reaches 95%. The parameter $s50$ is unconstrained, and $s95$ is
constrained to be larger than $s50$.

@@

@@

* $\omega_s$ are included if `spatial = 'on'` and omitted if `spatial = 'off'`.

@@

Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) but can be set to IID (`'iid'`; default), AR1 (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.

@@

@@

* $\epsilon_{\s,t}$ are \ldots

@@

Dropping other components for simplicity:

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$
where $\rho$ is the correlation between subsequent spatiotemporal random fields. The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance. The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is generally expected to be $-1 < \rho < 1$.

@@

Dropping other components for simplicity:

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_{t-1},  \:
\boldsymbol{\epsilon}_{t-1} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

where the distribution of the spatiotemporal field in the initial time step is the same as for the AR(1) model, but the absence of the $\rho$ parameter allows the spatiotemporal field to be non-stationary in time.

@@

@@

@@

where $\gamma_t$ is an example optional time-varying regression parameter and
$x_{\s,t}$ is the corresponding predictor value. This is defined via the
`time_varying` argument. E.g., `~ 0 + x`. The first value is given an implicit
uniform prior. I.e., the same variable should not appear in the fixed effect
formula since the initial value is estimated as part of the time-varying formula.

If `time` is defined in the call to `sdmTMB()`, then `time_varying ~ 1` implicitly represents a time-varying intercept and the intercept should be omitted from the main effects (`~ + 0` or `~ -1`).

@@

Offset terms can be included with the reserved word `offset` in the formula.
E.g., `y ~ x + offset`.

These are included in the linear predictor as

@@

>>>>

* Bold subscript $\s$ (a point in space) since it represents a vector of length 2

@@

* $\phi$ for all 'dispersion' parameters for consistency with the code

```{r, results='asis'}
symbols <- tibble::tribble(
  ~Symbol, ~Description,
  "$\\phi$", "A dispersion parameter for a distribution",
  "$\\mu$", "Mean",
  "$y$", "Observed response data",
  "$f$", "Link function",
  "$\\omega_{\\s}$", "Spatial random field at point $\\s$",
  "$\\zeta_{\\s, t}$", "Spatially varying coefficient random field at point $\\s$ and time $t$",
  "$\\epsilon_{\\s, t}$", "Spatiotemporal random field",
  "$\\boldsymbol{\\Sigma}_\\omega$", "Spatial random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\epsilon$", "Spatiotemporal random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\zeta$", "Spatially varying coefficient random field covariance matrix",
  "$\\boldsymbol{\\beta}$", "Parameter vector",
  "$\\boldsymbol{X}$", "Model predictor matrix",
  "$\\alpha_g$", "IID random intercept deviations",
  "$\\sigma_\\alpha^2$", "IID random intercept variance",
  "$\\delta_{\\s,t}$", "AR1 or random walk spatiotemporal deviations",
  "$\\rho$", "Correlation between random fields in subsequent time steps"
)
knitr::kable(symbols, caption = "Symbol notation and descriptions", escape = FALSE, format = "latex", booktabs = TRUE)
```

@@

* $y$ represents the response data;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}_{1}$, $\bm{X}_{2}$, and $\bm{X}_{3}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

@@

## Main effects

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \bm{X}_{1,\bm{s},t} \bm{\beta} \ldots \right)
\end{aligned}
$$

Within `sdmTMB()`, $\bm{X}_{1,\bm{s},t} \bm{\beta}$ is defined by the `formula` argument and represents the main-effect model matrix and a corresponding vector of coefficients. This main effect formula can contain optional penalized smoothers or non-linear functions as defined below.

@@

Smoothers in sdmTMB are generally implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`. 
Within these smooths, the same syntax commonly used in `mgcv::s()` can be applied, e.g. 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; the basis function dimensions may be specified, e.g. `+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, e.g. `+ s(month, bs = "cc", k = 12)`.

While mgcv can fit unpenalized (e.g., B-splines) or penalized splines (P-splines), sdmTMB only implements penalized splines. 
The penalized splines are constructed in sdmTMB using the function `mgcv::smooth2random()`, which transforms splines into random effects (and associated design matrices) that are estimable in a mixed-effects modelling framework.
This is a similar approach implemented in `gamm4` and `brms`, among other packages.

@@

The linear breakpoint or hockey stick model can be used to describe threshold or asymptotic responses. This function consists of two pieces, so that for $x < b_{1}$, $s(x) = x \cdot b_{0}$, and for $x > b_{1}$, $s(x) = b_{1} \cdot b_{0}$.
In both cases, $b_{0}$ represents the slope of the function up to a threshold, and the product $b_{1} \cdot b_{0}$ represents the value at the asymptote. No constraints are placed on parameters $b_{0}$ or $b_{1}$.

@@

Models with logistic threshold relationships between a predictor and the response can be fit with the form

@@

where $s$ represents the logistic function, 
$\psi$ is a scaling parameter (controlling the height of the y-axis for
the response, and is unconstrained), $\tau$ is an intercept, $s50$ is
a parameter controlling the point at which the function reaches 50% of the
maximum ($\psi$), and $s95$ is a parameter controlling the point at which the
function reaches 95%. The parameter $s50$ is unconstrained, and $s95$ is
constrained to be larger than $s50$.

@@

Spatial random fields, $\omega_s$, are included if `spatial = 'on'` and omitted if `spatial = 'off'`.

@@

@@

Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) and can be set to IID (`'iid'`; default), AR1 (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.

@@

IID spatiotemporal random fields (`spatiotemporal = 'iid'`) can be represented as

@@

where $\epsilon_{\s,t}$ represent random field deviations at point $\s$ and time $t$.
The random fields are assumed independent across time slices.

@@

AR1 spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time slice to the next. The are defined as

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \delta_{\s,t} \ldots \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$
where $\rho$ is the correlation between subsequent spatiotemporal random fields. The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.
The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is constrained to be $-1 < \rho < 1$.

@@

Random walk spatiotemporal random fields (`spatiotemporal = 'rw'`) represent a model where the difference in spatiotemporal deviations from one time slice to the next are IID. There defined as

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \delta_{\s,t} + \ldots \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_{t-1},  \:
\boldsymbol{\epsilon}_{t-1} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

where the distribution of the spatiotemporal field in the initial time step is the same as for the AR(1) model, but the absence of the $\rho$ parameter allows the spatiotemporal field to be non-stationary in time.
Note that, in contrast to the AR1 parametrization, the variance is no longer the steady-state marginal variance.

@@

Spatially varying coefficient (SVC) models are defined as

@@

Parameters can be modelled as time varying (a random walk) according to the form

@@

where $\gamma_t$ is an example optional time-varying regression parameter and $x_{\s,t}$ is the corresponding predictor value.
This is defined via the `time_varying` argument, assuming that the `time` argument is also supplied a column name.
For example, `time_varying ~ 0 + x`.
The first value is given an implicit uniform prior. 
I.e., the same variable should not appear in the fixed effect formula since the initial value is estimated as part of the time-varying formula.
The formula `time_varying = ~ 1` implicitly represents a time-varying intercept and the intercept should be omitted from the main effects (`~ + 0` or `~ -1`).

@@

Offset terms can be included with the reserved word `offset` in the formula.
E.g., `y ~ x + offset`.
These are included in the linear predictor as

@@

## [Censored-poisson] TODO
